---
title: "Assignment 1.2"
author: "Veera Koppula"
date: "12/4/2021"
output:
  pdf_document: default
---

```{r eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The first assignment is a review of graphical and data anlalysis. This purpose of this assignment is to provide a refresher of R and/or Python. The assignment is divided into three sections. Using R and/or Python, complete the following steps.

### 1. Import, Plot, Summarize, and Save Data
###Using the US Bureau of Labor Statistics data, choose a dataset that interests you. Then generate summary statistics for 2 variables, plot some of the features (e.g., histograms, box plots, density plots, etc.) of several variables, and save the data locally as CSV files.

#### Import the Data saved from BLS data store
```{r eval=TRUE, echo=FALSE}
setwd("/Users/veerareddykoppula/Downloads/Desktop/BRUIN/630/Assignment/Week1")
library('magrittr') 

men <- 'data/SeriesReport-20211204141549_39e88e.xlsx'
women <- 'data/SeriesReport-20211204141601_b6a497.xlsx'

# function to read and prepare data for analysis
preparetext <- function(path, index, datarow){

  # read excel data file
  df <- xlsx::read.xlsx(path,
                        sheetIndex = index,
                        stringsAsFactors = FALSE)
  
  # store value for gender (to be added later)
  gender <- df %>% 
    dplyr::filter(.[,1]=='Gender:') %>% 
    as.character() %>% 
    .[2]
  
  # store value for type (to be added later)
  type <- df %>% 
    dplyr::filter(.[,1]=='Type of estimate:') %>% 
    as.character() %>% 
    .[2]
  
  # store value for activity (to be added later)
  activity <- df %>% 
    dplyr::filter(.[,1]=='Activity:') %>% 
    as.character() %>% 
    .[2]
  
  # remove redundant rows
  df <- df %>% 
    dplyr::slice(datarow:nrow(.))
  
  # set first row as column name and remove first row
  colnames(df) <- as.character(unlist(df[1,]))
  df <- df[-1, ]
  
  # add subsequent data
  df <- df %>% 
    dplyr::mutate(Gender = gender,
                  Type = type,
                  Activity = activity)
  
  return(df)

}

```

#### merging the two data files to create one file for Household Time Use data as timeuse
```{r eval=TRUE, echo=FALSE}
timeuse <- preparetext(men, 1, 11) %>% 
  dplyr::bind_rows(preparetext(women, 1, 11)) %>% 
  dplyr::mutate(Estimate = as.numeric(Estimate),
                `Standard Error` = as.numeric(`Standard Error`))
```

#### Validating structure of Data
```{r eval=TRUE, echo=FALSE}
str(timeuse)
```


#### Generating summary statistics for the variables, I will print and add the details at the end of Step3 for the assignemnt
```{r eval=TRUE, echo=FALSE}
# Suppressing unnecessary warnings
options(warn=-1)

# Check data quality and basic statistics
stats  <- psych::describe(timeuse)

# Check detail descriptive statistics
stats_detail  <- Hmisc::describe(timeuse)
```

Based on the data review there are 2 continuous variables, Estimate and Standard Error. Rest of the 5 variables are discrete. I would like to create some basic plots for the 2 continous variables.
```{r eval=TRUE, echo=FALSE}
# set output figure size
options(repr.plot.width = 4, repr.plot.height = 3)

# Boxplot of the two continuous variables
DataExplorer::plot_boxplot(timeuse, by = 'Estimate')

# Histogram of the continuous variables (Estimate and Standard Error)
DataExplorer::plot_histogram(timeuse)

# density plot of estimate
par(mar=c(0,1,2,3))
plot(density(timeuse$Estimate) , main='Density of Estimates' , xlab="", ylim=c(0,2) , xaxt="n", las=1 , col="slateblue1" , lwd=4 )
```


### 2. Explore Some Bivariate Relations
###Use the same dataset within the same website to explore some bivariate relations (e.g. bivariate plot, correlation, table cross table etc.)

```{r eval=TRUE, echo=FALSE}
# explore bivariate relationship using cross table
gmodels::CrossTable(timeuse$Estimate, timeuse$`Standard Error`, expected = TRUE)
```
 
150.5167 is a very high Chi value, meaning the data (Estimate and Standard Error) does not fit very well.

The p-value is relatively large which indicates weak evidence against the null hypothesis, so we fail to reject the null hypothesis.

Now let us take a look at the descriptive statistics of the erntire dataset stored earlier.

### 3. Organize a Data Report
#### Generate a summary report. Make sure to include: summary for every variable, structure and type of data elements, discuss four results of your data.

```{r eval=TRUE, echo=FALSE}
#Printing stats details
print(stats_detail)
```
### Analysis:

Year - We have data between 2009-2019 and two instances per each year  for Men/Women.

Period - The period is a single value within tha data set.

Estimate - The estimate has somewhat even distribution which is not surprising given the data has men and women in equal proportion.

Standard Error - Standard Error picks between 0.033, 0.034,0.035 and 0.040.

Gender - The dataset has equal weightage on gender.

Since the summary descriptive statistics (stats) is generated as a data frame, I would output that in an excel file. But first I want to check what is in it.


```{r eval=TRUE, echo=FALSE}
# Printing stats
print(stats)
```

The variable names are converted into row names. I want to assign them to the first column of the dataf frame.

```{r eval=TRUE, echo=FALSE}
# Printing stats
# Convert the row name to first column
stats$variable <- rownames(stats)

# Reorder the columns to show variable as the first column
stats <- stats[c(14,1:13)]

# Convert infinity to NA
stats <- do.call(data.frame,lapply(stats, function(x) replace(x, is.infinite(x),NA)))

print(stats)
```
#### Saving the summary output into a XLS

```{r eval=TRUE, echo=FALSE}
# save the descriptive statistics in an excel file
library("xlsx")
xlsx::write.xlsx(stats, 
                 'output/Assignment 1.2.xlsx', 
                 sheetName = 'Descriptive Statistics',
                 row.names = FALSE,
                 append = TRUE)
```
